---
title: "Exploring Chi-Square Residuals with chi2residuals"
author: "Christophe Desterke"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring Chi-Square Residuals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

The chi2residuals package provides tools to explore and visualize chi-square residuals between two qualitative variables. It helps identify significant associations and deviations from expected frequencies using intuitive plots.

## Installation

```{r}
devtools::install_github("cdesterke/chi2residuals")
library(chi2residuals)
```

## Example Dataset


```{r}
data(patients)
str(patients)

```

We’ll explore the relationship between AgeGroup and PrimarySymptom.

## Preprocessing

```{r}
sub <- preProcess(patients, v1 = "AgeGroup", v2 = "PrimarySymptom")
head(sub)

```

## Computing Residuals

```{r}
residuals <- computeResiduals(sub, col1 = "AgeGroup", col2 = "PrimarySymptom")

```

## Heatmap Visualization

```{r resplot, fig.width=9, fig.height=4.5}
plotResiduals(
  residuals,
  col1 = "AgeGroup",
  col2 = "PrimarySymptom",
  themeSize = 18,
  labelSize = 3.5,
  colorLow = "turquoise",
  colorHigh = "purple",
  colorLabels = "white",
  title = "Significant residuals (p < 0.05)"
)

```


## Network Visualization

```{r netplot, fig.width=10, fig.height=5}
plotNet(
  residuals,
  var1 = "AgeGroup",
  var2 = "PrimarySymptom",
  node_colors = c(AgeGroup = "green", PrimarySymptom = "skyblue"),
  edge_colors = c(positive = "red", negative = "blue", nonsignificant = "lightgrey")
)

```

## Interpretation

Chi-square residuals highlight deviations between observed and expected frequencies:

🔴 Positive residuals: overrepresentation

🔵 Negative residuals: underrepresentation

⚪ Non-significant: consistent with independence

These visualizations help quickly identify strong associations between categories.

## Conclusion

chi2residuals is a practical tool for exploratory analysis of categorical data. It simplifies the detection of meaningful relationships and enhances interpretability through visual summaries.
